Introducing Statistical Design of Experiments to SPARQL Endpoint Evaluation

### 2013-05-06 exp1/iswc2013.tex
start of full explanation
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@514 43ef1190-90fd-4dd7-a769-d0121f02b80e

Then, we test some language features: 
``BGPComp'', which is a Basic Graph Pattern of varying complexity. The
first level is:
\verbatiminput{experiment/BGPComp-1}
and the second level is 
\verbatiminput{experiment/BGPComp-1}

@@

@@

>>>>

Then, we test some language features: 
``BGPComp'', which is a Basic Graph Pattern of varying complexity. The
first level is:
\verbatiminput{experiment/BGPComp-1}
and the second level is 
\verbatiminput{experiment/BGPComp-2}

@@

The code will be published\todo{github URL}.

@@

In the full factorial experiment, all combinations of the 8 factors
are executed, in total 256 runs.\todo{concrete example?} This is
called a $2^8$ factorial experiment, and we can see that the number of
runs is grows exponentially with the number of factors, thus making
large experiments prohibitively expensive. We shall nevertheless find
the following small example instructive. Note that each combination is
executed only once (i.e. it is \emph{unreplicated}, however, so even a
full factorial experiment compares well to a typical benchmark in
which each microbenchmark must be executed several times to obtain
enough data to compute a mean with reasonable accuracy.

The above experimental setup is executed across a design matrix
generated with the DoE.base package, which returns a data frame
consisting of the factors with a column containing the corresponding
curl measurements. To analyze the experiment, we first generate a
normal plot, see figure~\ref{fig:fullnormal}. In this plot, if the
null hypothesis $H_0$ holds, all estimated effects will fall on a
straight line, as their means would be zero. Any significant departure
from a straight line can be interpreted as a significant effect (see
Section~4.8 of \cite{wu2009experiments} for a detailed
explanation). In our case, it is most important to note that any
effect that is negative (i.e. has a negative slope) means the effect
enhances the performance of the SPARQL endpoint, the runtime
decreases. If ``Implement'' and the sum of its interactions are
negative, then it supports the alternative hypothesis, i.e. we have
successfully improved the performance.

To proceed beyond a visual judgement of significance, we may use
Lenth's method\todo{ref Lenth}, which is a simple method for
estimating significance in an experiment with no replicates. We plot
the results in figure~\ref{fig:fulllenth}. Visually, this is not as
useful, since the large number\todo{really include?} We see that we
have very significant effects. We also note that ``Implement'' itself
is significantly negative. We also note that the ``Implement:Lang''
interaction is significantly positive, as expected from our setup,
where ``Lang'' was the detrimental side-effect we simulated.

To investigate whether the detrimental side-effects cancels the
positive effect of our simulated optimization, we have to formulate a
hypothesis test. By inspecting the normal plot in figure~\ref{fig:fullnormal}
